---
title: "Automated gating of a PBMC panel in R"
subtitle: "Compensation, transformation and an openCyto template on eight samples"
author: "William Guesdon"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    theme: cosmo
    fig-width: 9
    fig-height: 6
    fig-dpi: 600
    grid:
      # Quarto's article layout gives the body 800px and puts 250px of margin on
      # each side. A report whose point is its figures needs that space, so the
      # body is wider and every figure chunk carries column: page, which spans
      # the body and both margins.
      sidebar-width: 250px
      body-width: 900px
      margin-width: 300px
      gutter-width: 1.5rem
knitr:
  opts_chunk:
    # knitr writes a display width from the file's own resolution, so raising
    # the resolution made every figure smaller on the page rather than sharper.
    # The figure fills its column instead, and the extra pixels go into detail.
    out.width: "100%"
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(flowCore)
  library(flowWorkspace)
  library(openCyto)
  library(ggcyto)
  library(ggplot2)
  library(knitr)
})

kOutputDir <- file.path("..", "output", "automated_gating_pbmc")

ReadOutput <- function(name) {
  path <- file.path(kOutputDir, name)
  if (!file.exists(path)) {
    stop(
      "The analysis output is missing: ", path, "\n",
      "Run scripts/01_automated_gating_pbmc.R first."
    )
  }
  read.csv(path, check.names = FALSE)
}

OutputImage <- function(name) {
  file.path(kOutputDir, name)
}
```

## Introduction

The FlowJo Basic Tutorial provides eight FCS files acquired on an LSRII from two
donors under four stimulation conditions. The files carry a compensation matrix
in the `$SPILLOVER` keyword and a viability dye channel.

This report applied compensation, estimated one logicle transform on the first
sample and applied it to all eight, and gated every sample with one openCyto
template. It reports the frequency of each population and the spread of those
frequencies across the eight samples.

### The gate hierarchy as a CSV

The gate hierarchy is stored in `gating/pbmc_gating_template.csv`. A template is
a text file that a version control system can track and a reviewer can read in a
pull request.

Maecker and colleagues sent pre-stained cells to 15 experienced laboratories.
Local gating gave a mean CV of 20.5 percent, and central gating of the same raw
files gave 4 percent [1]. The template makes central gating practical because
the analysis moves instead of the analyst.

## Methods

### The dataset

The design has two donors and four stimulation conditions, where `NS` means not
stimulated and `PI` means stimulated with PMA and ionomycin. The files were
acquired on an LSRII on 18 January 2013, and they come from the FlowJo Basic
Tutorial.

### Preprocessing

These files carry the matrix in the FCS `$SPILLOVER` keyword, which the
acquisition software wrote from single stained controls, and the analysis reads
that matrix and applies it. One transform is estimated on the first sample and
applied to all eight, because a per sample estimate would move a gate between
samples for a reason that has nothing to do with the biology.

### Gating

The template fixes the method and the channels, but it does not fix the cut
point. openCyto fits every gate on every sample separately, so a sample whose
distribution sits slightly differently still gets its own boundary. The
`nonDebris` row says to take `root` and, on `FSC-A` and `SSC-A`, fit a
`flowClust` gate and keep the positive population.

### Statistics

The coefficient of variation holds the biological difference between two donors,
the difference between four stimulation conditions, and the technical spread of
the assay. It holds no analyst variation, because one template gated every
sample. The pooled CV mixes the stimulation effect into the spread, and the
split CV does not, so it is closer to a measure of the assay itself.

### Software

The analysis used flowCore, flowWorkspace, openCyto, ggcyto, ggplot2 and knitr.
Every version is pinned in the container.

## Results

### The files and the panel

```{r}
#| label: sample-sheet

sample_sheet <- ReadOutput("sample_sheet.csv")
kable(
  sample_sheet[, c("file_name", "donor", "condition", "well", "stimulated")],
  caption = "The eight files, with the design parsed from the file names."
)
```

```{r}
#| label: event-counts

event_counts <- ReadOutput("event_counts.csv")
kable(event_counts, caption = "Events acquired per sample.")
```

#### The panel

```{r}
#| label: panel

panel <- ReadOutput("panel.csv")
kable(panel, caption = "Detectors and markers. A row with no marker is scatter or time.")
```

Two properties of this panel decide what the analysis can do. There is a
viability dye, so the analysis excludes dead cells instead of keeping them, and
there is `FSC-H` beside `FSC-A`, so the analysis excludes doublets. A panel with
only `FSC-A` cannot gate singlets at all, and a doublet then counts as a cell in
every population below it.

### Compensation

Spillover is the light from one fluorochrome that a second detector also reads.
The relationship is linear, so the correction multiplies the measured values by
the inverse of the spillover matrix.

```{r}
#| label: spillover-top

spillover_top <- ReadOutput("spillover_top10.csv")
spillover_top$percent <- round(100 * spillover_top$spill, 2)
kable(
  spillover_top[, c("from", "to", "percent")],
  caption = "The ten largest spillover values, as a percentage.",
  col.names = c("From detector", "Into detector", "Spillover, percent")
)
```

The largest value names the pair of detectors that compensation actually moves.

```{r}
#| label: spillover-matrix

spillover <- as.matrix(read.csv(
  file.path(kOutputDir, "spillover_matrix.csv"),
  row.names = 1, check.names = FALSE
))
kable(
  round(spillover, 3),
  caption = "The full spillover matrix. The diagonal is 1 by construction."
)
```

#### Doing this without a stored matrix

When the file carries no matrix, compute one from single stained controls with
`flowStats::spillover_ng()`. In Python the equivalent is a `flowkit.Matrix`
built from the controls. The common mistake is to compensate with a matrix from
a different acquisition day, and that mistake produces a population that does
not exist.

### Transformation

Fluorescence data spans several orders of magnitude, and after compensation it
holds negative values, which a log transform cannot represent. The logicle
transform is linear near zero and logarithmic above it, so the negative
population stays visible and the positive population stays separated.

```{r}
#| label: transform-channels

channels <- readLines(file.path(kOutputDir, "transformed_channels.txt"))
cat("Transformed channels:\n")
cat(paste0("  ", channels, collapse = "\n"))
```

### The gating template

```{r}
#| label: template

template <- read.csv(
  file.path("..", "gating", "pbmc_gating_template.csv"),
  check.names = FALSE
)
kable(
  template[, c("alias", "pop", "parent", "dims", "gating_method")],
  caption = "The gating template. Each row is one population."
)
```

```{r}
#| label: gating-tree
#| column: page
#| fig-cap: "The gate hierarchy that the template produced."

knitr::include_graphics(OutputImage("gating_tree.svg"))
```

### The gates

Each figure shows one gate on the first sample, and the fitted boundary is drawn
on the data it was fitted to.

```{r}
#| label: gate-figures
#| column: page
#| results: asis

populations <- readLines(file.path(kOutputDir, "population_paths.txt"))
populations <- setdiff(populations, "root")

for (population in populations) {
  image_name <- paste0(
    "gate_", gsub("[^A-Za-z0-9]+", "_", population), ".png"
  )
  image_path <- OutputImage(image_name)
  if (file.exists(image_path)) {
    cat("\n\n### ", population, "\n\n", sep = "")
    cat("![](", image_path, ")\n\n", sep = "")
  }
}
```

### Frequency of every population

```{r}
#| label: population-stats

population_stats <- ReadOutput("population_stats.csv")
kable(
  population_stats[, c("sample", "population", "count", "percent_of_parent")],
  digits = 2,
  caption = "Event count and percentage of the parent population, per sample."
)
```

```{r}
#| label: frequency-plot
#| column: page
#| fig-cap: "Population frequency by condition. One template gated all eight samples."

knitr::include_graphics(OutputImage("population_frequencies.svg"))
```

### Spread across the eight samples

```{r}
#| label: spread

spread <- ReadOutput("population_spread.csv")
kable(
  spread,
  digits = 2,
  caption = "Mean, standard deviation and coefficient of variation of each population frequency across all eight samples."
)
```

McNeil and colleagues measured this directly. 110 laboratories gated one shared
set of files, they produced 110 different gating approaches, and the standard
deviations were 2.5 to 25 times larger under laboratory specific gating than
under a consensus protocol [2].

```{r}
#| label: spread-by-condition

spread_by_condition <- ReadOutput("population_spread_by_condition.csv")
kable(
  spread_by_condition,
  digits = 2,
  caption = "The same statistics, split by stimulation condition. Splitting removes the between condition difference from the spread."
)
```

## Data and code availability

```bash
# Pull the dataset, about 207 MB of the 103 GB archive
./sync.sh pull datasets/flowjo

# Build the container. Every package version is pinned and asserted at build time.
./containers/build.sh

# Run the analysis
podman run --rm -v "$PWD:/work:z" -w /work everything-flow-cytometry:latest \
  Rscript scripts/01_automated_gating_pbmc.R

# Render this report
podman run --rm -v "$PWD:/work:z" -w /work everything-flow-cytometry:latest \
  quarto render reports/automated_gating_pbmc.qmd
```

```{r}
#| label: session-info

sessionInfo()
```

## References

[1] Maecker HT, McCoy JP Jr, for the FOCIS Human Immunophenotyping Consortium. A
model for harmonizing flow cytometry in clinical trials. Nat Immunol
2010;11(11):975-978. PMID 20959798.

[2] McNeil LK, et al. A harmonized approach to intracellular cytokine staining
gating. Cytometry A 2013;83A:728-38. PMID 23788464.

[3] Finak G, Frelinger J, Jiang W, Newell EW, Ramey J, Davis MM, Kalams SA, De
Rosa SC, Gottardo R. OpenCyto: an open source infrastructure for scalable,
robust, reproducible, and automated, end-to-end flow cytometry data analysis.
PLoS Comput Biol 2014;10(8):e1003806. PMID 25167361.
doi:10.1371/journal.pcbi.1003806.

[4] Van Gassen S, et al. FlowSOM: Using self-organizing maps for visualization and
interpretation of cytometry data. Cytometry A 2015. PMID 25573116.
