---
title: "Sample classification in the FlowCAP II challenges"
subtitle: "The FlowCAP II sample classification challenges, FR-FCM-ZZZV and FR-FCM-ZZZU"
author: "William Guesdon"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    theme: cosmo
    fig-width: 9
    fig-height: 6
    fig-dpi: 600
    grid:
      # Quarto's article layout gives the body 800px and puts 250px of margin on
      # each side. A report whose point is its figures needs that space, so the
      # body is wider and every figure chunk carries column: page, which spans
      # the body and both margins.
      sidebar-width: 250px
      body-width: 900px
      margin-width: 300px
      gutter-width: 1.5rem
knitr:
  opts_chunk:
    # knitr writes a display width from the file's own resolution, so raising
    # the resolution made every figure smaller on the page rather than sharper.
    # The figure fills its column instead, and the extra pixels go into detail.
    out.width: "100%"
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(ggplot2)
  library(knitr)
})

kOutputDir <- file.path("..", "output", "flowcap2")

ReadOutput <- function(name) {
  path <- file.path(kOutputDir, name)
  if (!file.exists(path)) {
    stop(
      "The analysis output is missing: ", path, "\n",
      "Run scripts/12_flowcap2_challenges.R first. ",
      "See the section 'Reproducing this'."
    )
  }
  read.csv(path, check.names = FALSE)
}

OutputImage <- function(name) {
  file.path(kOutputDir, name)
}

verdicts <- ReadOutput("claim_verdicts.csv")
scores <- ReadOutput("scores.csv")
```

## Introduction

The FlowCAP II benchmark tested whether computational pipelines can read class
labels from FCS files. Each entrant received half of the labels for training and
was scored on the held back half [1]. Two of the three challenge datasets are in
this archive, and both carry the labels and the split that the benchmark used.

This report applies a single feature rule to both deposits and scores the
results on the held out half.

```{r}
#| label: hvtn-design

kable(
  ReadOutput("hvtn_design.csv"),
  caption = "Challenge 3, the HVTN deposit. Each of 48 subjects contributes five files.",
  col.names = c("Treatment", "Split", "Files")
)
```

```{r}
#| label: heu-design

heu <- ReadOutput("heu_design.csv")
kable(
  heu[heu$Freq > 0, ],
  caption = "Challenge 1, the HEUvsUE deposit. Each of 44 infants contributes seven files.",
  col.names = c("Condition", "Stimulation", "Files")
)
```

## Methods

### The dataset

The HVTN deposit (FR-FCM-ZZZV) contains 48 subjects, each contributing five
files. The HEUvsUE deposit (FR-FCM-ZZZU) contains 44 infants, each contributing
seven files. Both deposits carry label files and the training and testing split
used by the FlowCAP II benchmark.

### Preprocessing

Every channel is transformed with an arcsinh at a fixed cofactor. A logicle
transform estimated per file was not used because a threshold taken from a
subject's control must be applied to that subject's stimulated file, and the two
files must sit on one scale. The logicle estimator refuses some files of the
HVTN deposit at every decade count.

### Gating

The cytokine gate uses the unstimulated control of each subject. The threshold
is placed at the 99.9th percentile of the control and applied to the stimulated
file of the same subject. The HVTN deposit provides two negative controls per
subject. One sets the threshold and the other measures the background at that
threshold.

### Statistics

The rule scores every gated feature on the training half alone, selects the
single best feature by area under the curve, places a threshold on it, and
applies that threshold to the held out half. The FlowCAP scoring uses recall,
precision, accuracy, and F-measure.

### Software

The analysis uses flowCore, ggplot2, and knitr. Every version is pinned in the
container.

## Results

### The rule that enters the benchmark

#### The cytokine gate

An unstimulated T cell is negative for all four cytokines, so a two component
mixture splits the negative population down the middle. The first attempt did
exactly that, and it reported 90 percent of CD4 T cells as interleukin 2
positive. The threshold is placed on the control at its 99.9th percentile and
applied to the stimulated file of the same subject. The HVTN deposit gives each
subject two negative controls. One sets the threshold and the other measures the
background at that same threshold.

#### One transform for every file

A threshold taken from a subject's control is applied to that subject's
stimulated file, so the two files have to sit on one scale. Every channel is
transformed with one arcsinh at a fixed cofactor.

```{r}
#| label: failures

failures <- ReadOutput("hvtn_failures.csv")
kable(
  if (nrow(failures) == 0) {
    data.frame(Reason = "none", Files = 0)
  } else {
    as.data.frame(table(Reason = failures$reason))
  },
  caption = "Why a file of the HVTN deposit could not be gated, out of 240.",
  col.names = c("Reason", "Files")
)
```

The gating failures reduce the number of subjects available for the paired
analysis.

### The population the paper names

```{r}
#| label: env-gag

kable(
  ReadOutput("env_against_gag.csv"),
  digits = 4,
  caption = "Every cytokine of every T cell subset, Env against Gag, paired within a subject. The values are percentages above the control threshold with the background removed.",
  col.names = c("Feature", "Subjects", "Env median", "Gag median",
                "Subjects where Env is higher", "p")
)
```

```{r}
#| label: env-gag-figure
#| column: page
#| fig-cap: "Env against Gag on every feature. The values are background removed, so a value near zero means no response above the subject's own control."

knitr::include_graphics(OutputImage("env_against_gag.svg"))
```

The paper reports that manual gating identified the CD4 interleukin 2 population
as discriminative, with the Env stimulated samples systematically higher than
the Gag stimulated ones. The deposit says the same. CD4 interleukin 2 is higher
in Env in 37 of the 47 paired subjects, and the paired test gives p below 0.001.

Three other features separate the two antigens as well, and one of them does it
more strongly. CD8 interferon gamma is the most significant of the eight, and it
runs the other way, with the Gag samples higher. The paper names one population
because manual gating looked at one population. An automated sweep over all
eight finds that population, and it also finds the three the paper does not
mention.

### Challenge scores

```{r}
#| label: scores

kable(
  scores,
  digits = 3,
  caption = "The four measures the FlowCAP table reports, for the training half and the held out half of each challenge.",
  col.names = c("Challenge", "Half", "Recall", "Precision", "Accuracy",
                "F-measure", "Samples")
)
```

```{r}
#| label: scores-figure
#| column: page
#| fig-cap: "The two challenges side by side. The dashed line is the accuracy of a coin."

knitr::include_graphics(OutputImage("scores.svg"))
```

```{r}
#| label: selected

ranking <- ReadOutput("hvtn_feature_ranking.csv")
kable(
  ranking,
  digits = 3,
  caption = "Every feature, ranked by how far its training area under the curve sits from a half. The top row is the feature the rule selected.",
  col.names = c("Feature", "Training area under the curve", "Separation")
)
```

```{r}
#| label: selected-figure
#| column: page
#| fig-cap: "The selected feature, with the threshold that the training half fitted. The testing points were never seen when the threshold was placed."

knitr::include_graphics(OutputImage("hvtn_selected_feature.svg"))
```

On the HVTN challenge one feature classifies the held out half at an accuracy
near 0.8. Six named algorithms reached perfect scores. The rule selected CD4
interleukin 4 rather than CD4 interleukin 2. The paper reports that four of the
eight methods which named their features chose one containing CD4 interleukin 2.
The claim about the selected feature is recorded as not reproduced.

On the HEUvsUE challenge the same rule reaches an accuracy of 0.875 on the
training half and 0.538 on the half it had not seen.

```{r}
#| label: heu-ranking

kable(
  head(ReadOutput("heu_feature_ranking.csv"), 10),
  digits = 3,
  caption = "The ten features that separate the HEUvsUE training half best. None of them survives the move to the held out half.",
  col.names = c("Feature", "Training area under the curve", "Separation")
)
```

The training half reaches an accuracy of 0.875 and the held out half reaches
0.538. The paper concludes that these samples may be unclassifiable on the basis
of the markers used. This analysis reproduces the pattern and cannot settle that
question.

### Claim verdicts

```{r}
#| label: verdicts

kable(
  verdicts[, c("claim_id", "short_name", "expected", "observed", "verdict")],
  caption = "Every claim the paper makes about these two challenge datasets.",
  col.names = c("Claim", "What the paper says", "Expected", "Measured",
                "Verdict")
)
```

```{r}
#| label: verdict-counts

kable(
  as.data.frame(table(verdicts$verdict)),
  caption = "How the six claims came out.",
  col.names = c("Verdict", "Claims")
)
```

## Discussion

### Principal findings

The two opposite results of the benchmark both reproduce from the deposited
files, using one rule that was written without reference to either. The HVTN
samples are classifiable. A single gated feature, chosen on the training half,
holds an accuracy near 0.8 on samples it had never seen. The HEUvsUE samples are
not. The same rule finds a feature that separates the training half at an
accuracy of 0.875 and then falls to a coin on the held out half.

The population the paper names is discriminative. CD4 interleukin 2 is higher
after Env stimulation in 37 of 47 subjects.

### The gap between the rule and the benchmark

Six named algorithms reached an F-measure of 1.0 on the HVTN challenge, and this
rule reaches 0.77. The gap is the rule and not the data. One feature and one
threshold cannot combine evidence, and the eight features here disagree with
each other. CD4 interleukin 4 and CD4 interleukin 2 both favour Env while CD8
interferon gamma favours Gag, so a method that reads all three at once has
information that a method reading one of them does not.

The benchmark was won by combining features rather than by better gating. A
single feature rule cannot combine evidence from multiple features.

### Limitations

This report does not enter the benchmark properly. FlowCAP scored a submission
against its own held out labels under a protocol, and this report scores itself.
The AML challenge, which is the third dataset and the one the paper treats as
the practical use case, is not in this archive.

The HEUvsUE deposit carries no training and testing split, so this report makes
one by alternating individuals. That split is deterministic and stated, and it
is not the split the benchmark used. Fifteen of the 44 infants also lack a
complete feature vector, because a lineage gate did not resolve in at least one
of their seven files. That leaves a small test set and it widens every interval
in the challenge scores.

### Further work

A rule that combines features is the next step, and the deposit supports it. The
method should be fixed before looking at the held out half, exactly as this rule
did. The HVTN deposit also carries a positive control for every subject,
stimulated with staphylococcal enterotoxin B. It is gated here and it is not
used, and it is the natural check on whether a subject's cells responded at all.

## Data and code availability

```bash
# Pull the two accessions, about 8 GB of the 103 GB archive
./sync.sh pull datasets/flowrepository/FlowRepository_FR-FCM-ZZZV_files
./sync.sh pull datasets/flowrepository/FR-FCM-ZZZU

# Build the container. Every package version is pinned and asserted at build time.
./containers/build.sh

# The analysis
podman run --rm -v "$PWD:/work:z" -w /work everything-flow-cytometry:latest \
  Rscript scripts/12_flowcap2_challenges.R

# This report
podman run --rm -v "$PWD:/work:z" -w /work everything-flow-cytometry:latest \
  quarto render reports/flowcap2_challenges.qmd
```

Both deposits carry their own label file in `attachments/`, so the class of
every sample travels with the data. The claims live in
`gating/flowcap2_paper_claims.csv`, each with the sentence it came from.

```{r}
#| label: session-info

sessionInfo()
```

## References

[1] Aghaeepour N, Finak G, FlowCAP Consortium, DREAM Consortium, Hoos H, Mosmann
TR, Brinkman R, Gottardo R, Scheuermann RH. Critical assessment of automated
flow cytometry data analysis techniques. Nat Methods 2013;10(3):228-238. PMID
23396282. PMC3906045. doi:10.1038/nmeth.2365.

[2] Spearman P, Lally MA, Elizaga M, Montefiori D, Tomaras GD, McElrath MJ, Hural
J, De Rosa SC, et al. A trimeric, V2-deleted HIV-1 envelope glycoprotein vaccine
elicits potent neutralizing antibodies but limited breadth of neutralization in
human volunteers. J Infect Dis 2011;203(8):1165-1173. PMID 21451004. PMC3068023.
doi:10.1093/infdis/jiq175.

[3] Hahne F, LeMeur N, Brinkman RR, Ellis B, Haaland P, Sarkar D, Spidlen J,
Strain E, Gentleman R. flowCore: a Bioconductor package for high throughput flow
cytometry. BMC Bioinformatics 2009;10:106. PMID 19358741.
doi:10.1186/1471-2105-10-106.
