---
title: "Automated gating of a human bone marrow panel in R"
subtitle: "Human bone marrow by flow cytometry and by mass cytometry, FR-FCM-ZYQ9 and FR-FCM-ZYQB"
author: "William Guesdon"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    theme: cosmo
    fig-width: 9
    fig-height: 6
    fig-dpi: 300
knitr:
  opts_chunk:
    # knitr writes a display width from the file's own resolution, so raising
    # the resolution to 300 dpi made every figure smaller on the page. The
    # figure fills the column instead, and the extra pixels go into sharpness.
    out.width: "100%"
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(ggplot2)
  library(knitr)
})

kOutputDir <- file.path("..", "output", "oetjen2018")

ReadOutput <- function(name) {
  path <- file.path(kOutputDir, name)
  if (!file.exists(path)) {
    stop(
      "The analysis output is missing: ", path, "\n",
      "Run scripts/10_oetjen2018_bone_marrow.R first. ",
      "See the section 'Reproducing this'."
    )
  }
  read.csv(path, check.names = FALSE)
}

OutputImage <- function(name) {
  file.path(kOutputDir, name)
}

verdicts <- ReadOutput("claim_verdicts.csv")
```

## Introduction

Oetjen and colleagues profiled bone marrow from 20 healthy adults by single cell
RNA sequencing, by mass cytometry and by flow cytometry [1]. The flow files are
deposited in FR-FCM-ZYQ9 and the mass cytometry files are in FR-FCM-ZYQB. The
cohort spans 24 to 84 years of age. This report measures marker stability under
an automated density cut, agreement between flow and mass cytometry on T cell
subsets, and age trends in the memory compartments.

```{r}
#| label: coverage

kable(
  ReadOutput("panel_coverage.csv"),
  caption = "The flow deposit. Five stained panels and one unstained control for each sample.",
  col.names = c("Panel", "Files")
)
```

```{r}
#| label: cohort

kable(
  ReadOutput("cohort.csv"),
  caption = "The cohort, read from Table 1 of the paper into gating/oetjen2018_donor_metadata.csv."
)
```

Two of the samples are second aspirations from donors who were sampled twice,
which is why 22 samples come from 20 donors. The paper states a median age of 57
years. Table 1 gives a median of 52.5 years across the 20 donors and 55.5 years
across the 22 samples. The difference is recorded.

## Methods

### The dataset

Bone marrow samples from 20 healthy adults were profiled by flow cytometry and
by mass cytometry. The flow files are in FR-FCM-ZYQ9 and the mass cytometry
files are in FR-FCM-ZYQB. The cohort spans 24 to 84 years of age. Two donors
were sampled twice, which gives 22 samples.

### Preprocessing

Every file came from one LSRFortessa. The detector names are the same in all 132
files. A detector names a marker once the panel is known. The published panel
design supplies the fallback when a file carries no marker name. The count of
CD45 positive events is taken before the scatter gate, because the deposited
polygon is a lymphocyte gate. The mass cytometry files have no scatter and no
conventional viability dye. The two iridium DNA channels and the platinum
channel replace those steps. The transform is arcsinh with a cofactor of 5.

### Gating

The scatter gate and the viability gate are the depositors' own, extracted from
the cytoflow workflow file. The viability gate direction is read from the
deposited gate rather than from a convention. The `LiveMask()` function applies
this direction.

### Statistics

Marker stability is measured as the spread of percent positive across samples.
The automated cut uses a density minimum with a two component mixture as the
fallback. Technology agreement is measured by Spearman correlation. Age trends
are tested by Spearman correlation on the flow data.

### Software

The analysis uses flowCore, ggplot2 and knitr. Every version is pinned in the
container.

## Results

### The deposited viability gate

The deposit carries a cytoflow workflow with two polygons. The first is named
`lymphocytes` and it sits on forward scatter height against side scatter height.
The second is named `dead_lymphocytes` and it covers the low end of the V545
detector. The LIVE/DEAD channel is a fixable viability dye. The deposited gate
assigns live status to events above the polygon boundary.

```{r}
#| label: gates

gates <- read.csv(file.path("..", "gating", "oetjen2018_manual_gates.csv"),
                  check.names = FALSE)
kable(
  do.call(rbind, lapply(split(gates, gates$gate), function(piece) {
    data.frame(
      Gate = piece$gate[1],
      `X channel` = piece$x_channel[1],
      `Y channel` = piece$y_channel[1],
      Vertices = nrow(piece),
      `X range` = sprintf("%.0f to %.0f", min(piece$x), max(piece$x)),
      `Y range` = sprintf("%.0f to %.0f", min(piece$y), max(piece$y)),
      check.names = FALSE, stringsAsFactors = FALSE
    )
  })),
  caption = "The two polygons the depositors shipped, extracted into gating/oetjen2018_manual_gates.csv.",
  row.names = FALSE
)
```

On the V545 detector the `dead_lymphocytes` polygon reaches to about 2,553, and
the pipeline reported CD45 positive cells as 4 percent of live events in a bone
marrow sample.

```{r}
#| label: hierarchy

kable(
  ReadOutput("hierarchy.csv"),
  caption = "The median event count at each step, by panel. The scatter gate and the viability gate are the depositors' own.",
  col.names = c("Panel", "Files", "Acquired", "Singlets", "After the scatter gate",
                "Live", "Parent gate")
)
```

### Resolving thirteen colours across five panels

A large share of the files carry no marker name. The published panel design is
committed as `gating/oetjen2018_panels.csv`, and it supplies the fallback.

```{r}
#| label: resolution

kable(
  ReadOutput("channel_resolution_summary.csv"),
  caption = "How each channel was matched to a marker across the stained files.",
  col.names = c("Resolved by", "Channels")
)
```

The FCS header resolved most channels. The panel design supplied the remainder.

```{r}
#| label: events

kable(
  ReadOutput("event_summary.csv"),
  caption = "The event counts the paper reports, measured on the deposited files."
)
```

The paper reports a median of 196,000 CD45 positive events collected, with a
25th to 75th percentile of 100,000 to 278,000. The count here is taken before
the scatter gate.

### Marker stability under the automated cut

The upstream gates come from the depositors. The automated cut below them uses a
density minimum with a two component mixture as the fallback.

```{r}
#| label: stability

stability <- ReadOutput("marker_stability.csv")
kable(
  head(stability, 15),
  digits = 1,
  caption = "The fifteen least stable marker and panel pairs, ranked by the spread of the percent positive across the samples.",
  col.names = c("Panel", "Marker", "Samples", "Median percent", "Minimum",
                "Maximum", "Range", "CV")
)
```

The least stable pairs show ranges that exceed 80 percentage points.

```{r}
#| label: stability-verdict

kable(
  ReadOutput("stability_verdict.csv"),
  caption = "The spread of every marker and panel pair in the deposit."
)
```

```{r}
#| label: spread-figure
#| fig-cap: "Every marker of every panel, across the samples. A box that spans the whole axis is a cut the rule could not place."

knitr::include_graphics(OutputImage("marker_spread.svg"))
```

Most marker and panel pairs span more than 80 percentage points across samples.
The two claims that need a reliable lineage frequency from every panel, the
order of the major populations and which population is largest, are recorded as
not measurable by this method.

The automated density cut fails here because the marker distribution has no
reliable second mode. The tissue holds a developmental continuum.

### T cell subsets by two technologies

The T panel and the mass cytometry panel both carry CD3, CD4, CD8, CD45RA and
CCR7. Both produce the naive, central memory, effector memory and terminally
differentiated subsets.

```{r}
#| label: technology

agreement <- ReadOutput("technology_agreement.csv")
kable(
  agreement,
  digits = 3,
  caption = "The same measure by flow cytometry and by mass cytometry, on the donors that both technologies covered.",
  col.names = c("Measure", "Donors", "Flow median", "Mass median",
                "Spearman rho", "p")
)
```

The correlation does not reach significance for any subset.

```{r}
#| label: technology-figure
#| fig-cap: "Each point is one donor. The dashed line is equality."

knitr::include_graphics(OutputImage("flow_against_mass.svg"))
```

The paper reports a strong correlation between the two technologies on the CD4
and CD8 subsets. This analysis does not find one. The mass cytometry deposit
covers eight donors, the T panel did not gate every flow file, and the donors
that survive both routes are fewer still. A correlation on that many points has
low power. The CD45RA and CCR7 cuts are the least reliable in the panel. A
subset built from two unreliable cuts inherits both.

### The age trend and the repeated aspiration

```{r}
#| label: age

kable(
  ReadOutput("age_trend.csv"),
  digits = 3,
  caption = "The T cell subsets against donor age, by Spearman correlation on the flow data.",
  col.names = c("Measure", "Samples", "Spearman rho", "p")
)
```

The memory compartments show positive correlations with age.

```{r}
#| label: age-figure
#| fig-cap: "Each point is one sample. The line is a linear fit, drawn to show the direction."

knitr::include_graphics(OutputImage("age_trend.svg"))
```

The memory compartment rises with age in both the CD4 and the CD8 lineage. Both
correlations are significant. The terminally differentiated CD8 compartment
moves the other way. This report does not resolve the direction. The claim the
paper makes is about the memory population, and that claim reproduces from the
deposited files.

```{r}
#| label: repeats

repeats <- ReadOutput("repeated_aspiration.csv")
kable(
  repeats,
  digits = 1,
  caption = "The two aspirations of the donors who were sampled twice.",
  col.names = c("Donor", "Measure", "First aspiration", "Second aspiration",
                "Absolute difference")
)
```

The absolute differences between aspirations are small for all measures.

### Claim verdicts

```{r}
#| label: verdicts

kable(
  verdicts[, c("claim_id", "short_name", "expected", "observed", "verdict")],
  caption = "Every claim the paper makes from the flow or the mass cytometry data.",
  col.names = c("Claim", "What the paper says", "Expected", "Measured",
                "Verdict")
)
```

```{r}
#| label: verdict-counts

kable(
  as.data.frame(table(verdicts$verdict)),
  caption = "How the eight claims came out.",
  col.names = c("Verdict", "Claims")
)
```

## Discussion

### Principal findings

The deposited cytoflow file carries two polygons. One of them reverses an
assumption that a conventional pipeline makes. The age claim reproduces from the
deposited files with a significant correlation in both T cell lineages. The
analysis shares no software with the original. The concordance between two
aspirations of one donor also reproduces.

### Limitations

The analysis cannot test the paper claim about the correlation between mass
cytometry and flow cytometry. The shared donors are too few and the cuts are the
least reliable in the panel. The bone marrow populations are not ranked, because
the automated cut does not supply a reliable lineage frequency on this deposit.
The single cell RNA sequencing is not included. Those data are in GEO and not in
this archive, so the claim about a discrepancy between sequencing and flow
cytometry is out of reach.

### Further work

The authors' full gating strategy is in Supplemental Table 2 of the paper. It is
not in the deposit in a machine readable form. The lineage frequencies would
come from stated boundaries rather than from an automated rule. The unstained
control of each sample is in the deposit. This analysis does not use it.

## Data and code availability

```bash
# Pull the two accessions, about 17 GB of the 103 GB archive
./sync.sh pull datasets/flowrepository/FR-FCM-ZYQ9
./sync.sh pull datasets/flowrepository/FR-FCM-ZYQB

# Build the container. Every package version is pinned and asserted at build time.
./containers/build.sh

# The analysis
podman run --rm -v "$PWD:/work:z" -w /work everything-flow-cytometry:latest \
  Rscript scripts/10_oetjen2018_bone_marrow.R

# This report
podman run --rm -v "$PWD:/work:z" -w /work everything-flow-cytometry:latest \
  quarto render reports/oetjen2018_bone_marrow.qmd
```

The cohort table is derived from Table 1 of the paper and it is committed as
`gating/oetjen2018_donor_metadata.csv`. The panel design is derived from
Supplemental Table 1B and committed as `gating/oetjen2018_panels.csv`. The two
polygons are extracted from the deposited cytoflow workflow and committed as
`gating/oetjen2018_manual_gates.csv`.

```{r}
#| label: session-info

sessionInfo()
```

## References

[1] Oetjen KA, Lindblad KE, Goswami M, Gui G, Dagur PK, Lai C, Dillon LW, McCoy
JP, Hourigan CS. Human bone marrow assessment by single-cell RNA sequencing,
mass cytometry, and flow cytometry. JCI Insight 2018;3(23):e124928. PMID
30518681. PMC6328018. doi:10.1172/jci.insight.124928.

[2] Hahne F, LeMeur N, Brinkman RR, Ellis B, Haaland P, Sarkar D, Spidlen J,
Strain E, Gentleman R. flowCore: a Bioconductor package for high throughput flow
cytometry. BMC Bioinformatics 2009;10:106. PMID 19358741.
doi:10.1186/1471-2105-10-106.
